The Impact of Alternation

### 2017-05-08 draft.tex
proof of Thm. 1 added

We denote by $\fv{S}{\phi}$ the set of free (i.e.\ not under the scope
of an existential quantifier) variables of sort $S$ in $\phi$ and
write $\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
that $\sigma(t) = \sigma(x)$, let $\phi[t/x]$ be the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i=1,\ldots,n$.
%
%% A formula is in positive normal form (PNF) if negation occurs only
%% within literals and in disjunctive normal form (DNF) if it is a
%% disjunction of conjunctions of literals $\bigvee_{i=1}^N
%% \bigwedge_{j=1}^{M_i} \ell_{ij}$.
%
The size of a formula $\phi$, denoted $\len{\phi}$ is the number of
symbols occuring in it.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$. The
\emph{theory} $\theory(\sig,\I)$ is the set of satisfiable formulae
written in the signature $\sig$, with the interpretation $\I$.

@@

This definition generalizes the classical notion of words from a
finite alphabet to using a possibly infinite alphabet
$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
infinite, we can map the elements of $\Sigma$ into designated elements
of $\Data^\I$ and use a special variable to encode the input
events. However, keeping $\Sigma$ explicit in the following simplifies
several technical points below, without cluttering the presentation.

@@

An \emph{alternating data automaton} (or, simply, and automaton) is a
tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q$ is a finite set of variables of sort $\Bool$, also called
  \emph{states},
%
\item $\iota \in \Form^+(Q,\emptyset)$ is a formula describing the
  \emph{initial configurations},
%
\item $F \subseteq Q$ is a set of \emph{final} states, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
  \mid x \in \vec{x}\}$.
\end{compactitem}
Intuitively, in each formula describing a transition rule,
$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
values of the variables of $\A$. Observe further, that the initial
values of the variables are left unconstrained, because the initial
configuration does not contain free data variables. The size of $\A$
is defined as $\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
a time-stamped set of variables. For an input even $a \in \Sigma$ and
a formula $\phi$ we write $\Delta^k(\phi,a)$ for the formula obtained
from $\phi$ by simultaneously replacing each state $q \in
\fv{\Bool}{\phi}$ by the formula
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$. Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
and, for all $k > 0$, we have $\phi_k \equiv
\Delta^k(\phi_{k-1},a_k)$. We slightly abuse notation and write
$\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ in the following. We say that
$\A$ \emph{accepts} $w$ iff $\I,\nu \models \phi_n$, where $\nu$ is
some valuation that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$, and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words accepted by $\A$.

In this section we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, do there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is it the case that
  $L(\A) = \emptyset$ ?
\end{compactenum}

It is well known that other problems, such as \emph{universality}
(given an automaton $\A$ with variables $\vec{x}$, does $L(\A) =
\Sigma[\vec{x}]^*$) and \emph{inclusion} (given automata $\A_1$ and
$\A_2$ with the same set of variables, does $L(\A_1) \subseteq
L(\A_2)$ ?) can be reduced to the above problems. Observe furthermore
that we do not consider cases in which the sets of variables in the
two automata differ. An interesting problem in this case is: given
automata $\A_1$ and $\A_2$, with variables $\vec{x}_1$ and
$\vec{x}_2$, respectively, such that $\vec{x}_1 \subseteq \vec{x}_2$,
does $L(\A_1) \subseteq \proj{L(\A_2)}{\vec{x}_1}$, where
$\proj{L(\A_2)}{\vec{x}_1}$ is the projection of the set of words
$L(\A_2)$ onto the variables $\vec{x}_1$? This problem is considered
as future work.

@@

\begin{proposition}\label{prop:overline}
  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
  mapping each $q \in Q$ a value $\nu(q) \in \booli$ and each $x \in
  \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
  valuation that assigns each $q \in Q$ the value $\neg\nu(q)$ and
  each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
  $\I,\nu \models \phi \iff \I,\nu' \not\models \overline{\phi}$. 
\end{proposition}
\proof{Immediate, by induction on the structure of $\phi$. \qed}

In the following let $\A_1 = \tuple{\vec{x},Q_1,\iota_1,F_1,\Delta_1}$
and $\A_2 = \tuple{\vec{x},Q_2,\iota_2,F_2,\Delta_2}$, where
w.l.o.g. we assume that $Q_1 \cap Q_2 = \emptyset$. We define $\A_\cup
= \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \vee \iota_2,F_1\cup
  F_2,\Delta_1\cup\Delta_2}$, $\A_\cap = \tuple{\vec{x},Q_1\cup
  Q_2,\iota_1 \wedge \iota_2, F_1\cup F_2, \Delta_1\cup\Delta_2}$ and
$\overline{\A_1} = \langle \vec{x},Q_1,\iota_1,Q_1\setminus
F_1,\overline{\Delta_1} \rangle$, where $\overline{\Delta_1}(q,a)
\equiv \overline{\Delta_1(q,a)}$, for all $q \in Q_1$ and $a \in
\Sigma$. The following lemma shows the correctness of these
definitions:

@@

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
only atoms $x \geq k$ and $x = \overline{x} + k$, for $k \in \zed$,
with the classical interpretation of the function symbols on
integers. Since reachability of a control state is undecidable for
AVASS, the emptiness problem is undecidable for ADA.

Consequently, we give up on the guarantee for termination and develop
semi-algorithms that are supposed to be: \begin{compactenum}
\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
  the procedure will terminate and return a word $w \in L(\A)$), and
%
\item \emph{complete relative to termination}: if the procedure terminates
  without returning such a word, then $L(\A) = \emptyset$. 
\end{compactenum}

@@

We generalize the post-image function to finite sequences of input
events, as follows: $\Post{\A}(\phi,\varepsilon) \equiv \phi$ and
$\Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a)$, for any $u
\in \Sigma^*$. Then the emptiness problem for $\A$ becomes: does there
exist $u \in \Sigma^*$ such that the formula $\Accept{\A}(u) \equiv
\Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q
\rightarrow \bot)$ is satisfiable ?  Observe that, since we ask a
satisfiability query, the final states of $\A$ need not be
constrained\footnote{ Since each state occurs positively in
  $\Accept{\A}(u)$, this formula has a model iff it has a model with
  every $q \in F$ set to true.}. A na\"ive semi-algorithm enumerates
all finite sequences and checks the satisfiability of $\Accept{\A}(u)$
for each $u \in \Sigma^*$, using a decision procedure for the theory
$\theory(\sig,\I)$.

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
$\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$, thus
$\Accept{\A}(vw)$ is satisfiable as well, and there is no need for
$u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, in practice, the antichain-based semi-algorithm
diverges in a large number of cases, since, in many interpretations,
there exist infinite antichains, e.g.\ $q \wedge x \teq 0, q \wedge x
\teq 1, \ldots$ for $\Data^\I = \nat$.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation
$\nu : \vec{x} \mapsto \Data^\I$ of the data
variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu
  \not\models \Accept{\A}(u)$.
\end{compactenum}
A formula $\phi(Q,\vec{x})$ \emph{defines} $\Inv$ iff for all $\beta:
Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow \Data^\I$,
$\beta\cup\nu \models \phi \iff \nu \in \Inv(\beta)$. If $\Inv$
satisfies only the first point above, we call it simply an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota = \Post{\A}(\iota,\varepsilon)$, in which the
data variables are unconstrained, and over-approximates the set of
reachable valuations (\ref{it1:inv}), while excluding those valuations
which satisfy the acceptance condition (\ref{it2:inv}).

\begin{lemma}\label{lemma:safety-invariant}
  For any automaton $\A$, we have $L(\A)
= \emptyset$ iff $\A$ has a safety invariant. 
\end{lemma}
\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
  u \in \Sigma^* ~.~ \beta\cup\nu \models
  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
  straightforward. \qed}

@@

The solution we adopt is the use of abstraction to ensure termination
of invariant computations. However, it is worth pointing out from the
start that abstraction will allow us to build invariants which are not
necessarily safety invariants. To meet the latter condition, we resort
to counterexample guided abstraction refinement (CEGAR).

Formally, let $\preds$ be a finite set of formulae with variables from
$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
can occur negated in some $\phi \in \preds$. We assume that $\preds$
is fixed for now, and refer to these formulae as \emph{predicates}.
Given a formula $\phi$, we denote by $\abs{\phi} \equiv \bigwedge
\set{\pi \in \preds \mid \phi \models \pi}$ the abstraction of $\phi$
w.r.t. the predicates in $\preds$. The abstract version of the
post-image and acceptance condition are defined as follows:
\[\begin{array}{llcl}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi & 
\AbsPost{\A}(\phi,ua) & \equiv & \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
&
\AbsAccept{\A}(u) & \equiv & \AbsPost{\A}(\iota,u) \wedge\!\!\!\bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)
\end{array}\]
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=1}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \eqiff
  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}

>>>>

We denote by $\fv{S}{\phi}$ the set of free (i.e.\ not under the scope
of an existential quantifier) variables of sort $S$ in $\phi$ and
write $\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
that $\sigma(t) = \sigma(x)$, let $\phi[t/x]$ be the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i=1,\ldots,n$.
%
%% A formula is in positive normal form (PNF) if negation occurs only
%% within literals and in disjunctive normal form (DNF) if it is a
%% disjunction of conjunctions of literals $\bigvee_{i=1}^N
%% \bigwedge_{j=1}^{M_i} \ell_{ij}$.
%
The size of a formula $\phi$, denoted as $\len{\phi}$, is the number
of symbols occuring in it.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
in the signature $\sig$, with the interpretation $\I$.

@@

This definition generalizes the classical notion of words from a
finite alphabet to the possibly infinite alphabet
$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
infinite, we can map the elements of $\Sigma$ into designated elements
of $\Data^\I$ and use a special variable to encode the input
events. However, keeping $\Sigma$ explicit in the following simplifies
several technical points below, without cluttering the presentation.

@@

An \emph{alternating data automaton} (ADA) or, simply, an automaton,
is a tuple $\A =
\tuple{\vec{x},Q,\iota,F,\Delta}$, where: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q$ is a finite set of variables of sort $\Bool$, called
  \emph{states},
%
\item $\iota \in \Form^+(Q,\emptyset)$ is positive boolean combination
  of states defining the \emph{initial configurations},
%
\item $F \subseteq Q$ is a set of \emph{final states}, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
  \mid x \in \vec{x}\}$.
\end{compactitem}
Intuitively, in each formula $\Delta(q,a)$ describing a transition
rule, the variables $\overline{\vec{x}}$ track the previous and
$\vec{x}$ the current values of the variables of $\A$. Observe
further, that the initial values of the variables are left
unconstrained, as the initial configuration does not contain free
data variables. The size of $\A$ is defined as $\len{\A} = \len{\iota}
+ \sum_{(q,a) \in Q\times\Sigma}\len{\Delta(q,a)}$.

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
a set of time-stamped variables. For an input event $a \in \Sigma$ and
a formula $\phi$, we write $\Delta^k(\phi,a)$ for the formula obtained
from $\phi$ by simultaneously replacing each state $q \in
\fv{\Bool}{\phi}$ by the formula
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$. Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
and, for all $k > 0$, we have $\phi_k \equiv
\Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse notation and write
$\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
where $\nu$ is some valuation that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words accepted by $\A$.

In this paper we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, do there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is it the case that
  $L(\A) = \emptyset$ ?
\end{compactenum}

It is well known that other problems, such as \emph{universality}
(given automaton $\A$ with variables $\vec{x}$, does $L(\A) =
\Sigma[\vec{x}]^*$?) and \emph{inclusion} (given automata $\A_1$ and
$\A_2$ with the same set of variables, does $L(\A_1) \subseteq
L(\A_2)$?) can be reduced to the above problems. Observe furthermore
that we do not consider cases in which the sets of variables in the
two automata differ. An interesting problem in this case would be: given
automata $\A_1$ and $\A_2$, with variables $\vec{x}_1$ and
$\vec{x}_2$, respectively, such that $\vec{x}_1 \subseteq \vec{x}_2$,
does $L(\A_1) \subseteq \proj{L(\A_2)}{\vec{x}_1}$, where
$\proj{L(\A_2)}{\vec{x}_1}$ is the projection of the set of words
$L(\A_2)$ onto the variables $\vec{x}_1$? This problem is considered
as future work.

@@

\begin{proposition}\label{prop:overline}
  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
  mapping each $q \in Q$ to a value $\nu(q) \in \booli$ and each $x \in
  \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
  valuation that assigns each $q \in Q$ the value $\neg\nu(q)$ and
  each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
  $\I,\nu \models \phi$ if and only if $\I,\nu' \not\models \overline{\phi}$. 
\end{proposition}
\proof{Immediate, by induction on the structure of $\phi$. \qed}

In the following let $\A_i =
\tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, where
w.l.o.g. we assume that $Q_1 \cap Q_2 = \emptyset$. We define: 
\[\begin{array}{rcl}
\A_\cup & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \vee \iota_2,F_1\cup F_2,\Delta_1\cup\Delta_2} \\ 
\A_\cap & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \wedge \iota_2, F_1\cup F_2, \Delta_1\cup\Delta_2} \\ 
\overline{\A_1} & = & \langle \vec{x},Q_1,\iota_1,Q_1\setminus F_1,\overline{\Delta_1} \rangle
\end{array}\] 
where $\overline{\Delta_1}(q,a) \equiv \overline{\Delta_1(q,a)}$, for
all $q \in Q_1$ and $a \in \Sigma$. The following lemma shows the
correctness of the above definitions:

@@

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
only atoms $x \geq k$ and $x = \overline{x} + k$, for $k \in \zed$,
with the classical interpretation of the function symbols on
integers. Since reachability of a control state is undecidable for
AVASS \cite{LINCOLN92}, the emptiness problem is undecidable for ADA.

Consequently, we give up on the guarantee for termination and develop
semi-algorithms that are supposed to be: \begin{compactenum}
\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
  the procedure will terminate and return a word $w \in L(\A)$, and
%
\item \emph{complete relative to termination}: if the procedure terminates
  without returning such a word, then $L(\A) = \emptyset$. 
\end{compactenum}

@@

We generalize the post-image function to finite sequences of input
events, as follows, for any $u \in \Sigma^*$:
\[\begin{array}{l}
\Post{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm} \Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a) \\
\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)
\end{array}\]
Then the emptiness problem for $\A$ becomes: does there exist $u \in
\Sigma^*$ such that the formula $\Accept{\A}(u)$ is satisfiable?
Observe that, since we ask a satisfiability query, the final states of
$\A$ need not be constrained\footnote{ Since each state occurs
  positively in $\Accept{\A}(u)$, this formula has a model iff it has
  a model with every $q \in F$ set to true.}. A na\"ive semi-algorithm
enumerates all finite sequences and checks the satisfiability of
$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
for the theory $\theory(\sig,\I)$.

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
$\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$, thus
$\Accept{\A}(vw)$ is satisfiable as well, and there is no need for
$u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, the antichain-based semi-algorithm diverges in a large
number of cases, since infinite antichains exist in many
interpretations, e.g.\ $q \wedge x \teq 0,~ q \wedge x \teq 1, \ldots$
for $\Data^\I = \nat$.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
sequence $u \in \Sigma^*$ of input events, the following
hold: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
  \nu \not\models \Accept{\A}(u)$.
\end{compactenum}
A formula $\phi(Q,\vec{x})$ \emph{defines} $\Inv$ iff for all $\beta:
Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow \Data^\I$,
$\I,\beta\cup\nu \models \phi$ if and only if $\nu \in
\Inv(\beta)$. If $\Inv$ satisfies only the first point above, we call
it an \emph{invariant}. Intuitively, a safety invariant maps every
boolean valuation into a set of data valuations, that contains the
initial configuration $\iota = \Post{\A}(\iota,\varepsilon)$, whose
data variables are unconstrained, over-approximates the set of
reachable valuations (\ref{it1:inv}) and excludes the valuations
satisfying the acceptance condition (\ref{it2:inv}).

\begin{lemma}\label{lemma:safety-invariant}
  For any automaton $\A$, we have $L(\A) = \emptyset$ if and only if
  $\A$ has a safety invariant.
\end{lemma}
\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
  u \in \Sigma^* ~.~ \beta\cup\nu \models
  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
  straightforward. \qed}

@@

The solution we adopt is abstraction to ensure the termination of
invariant computations. However, it is worth pointing out from the
start that abstraction will allow us to build invariants which are not
necessarily safety invariants. To meet the latter condition, we resort
to counterexample guided abstraction refinement (CEGAR).

Formally, let $\preds$ be a finite set of formulae with variables from
$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
can now occur negated within the formulae $\phi \in \preds$. We fix
$\preds$ from now on and refer to these formulae as
\emph{predicates}. Given a formula $\phi$, we denote by $\abs{\phi}
\equiv \bigwedge \set{\pi \in \preds \mid \phi \models \pi}$ the
abstraction of $\phi$ w.r.t. the predicates in $\preds$. The abstract
version of the post-image and acceptance condition are defined as
follows, for any $u \in \Sigma^*$:
\[\begin{array}{l}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)
\end{array}\]
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=1}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \eqiff
  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}

We are left with fulfilling the second point from the definition of a
safety invariant. To this end, suppose that, for a given set $\preds$
of predicates, the invariant $\abs{\Inv}$, defined by the previous
lemma, meets point (\ref{it1:inv}) but not point (\ref{it2:inv}) from
the definition of invariants, where $\Post{\A}$ and $\Accept{\A}$
replace $\AbsPost{\A}$ and $\AbsAccept{\A}$, respectively. In
other words, there exists a finite sequence $u \in \Sigma^*$ such that
$\nu \in \abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models
\AbsAccept{\A}(u)$, for some boolean $\beta : Q \rightarrow \booli$
and data $\nu : \vec{x} \rightarrow \Data^\I$ valuations. Such
a $u\in\Sigma^*$ is called a \emph{counterexample}.
